\protect Connecting every bit of knowledge: \\ The structure of Wikipedia's first link network

### 2016-03-07 writeup/wikipedia-network.tex
chris edits incorporated up to results section

\title{\protect
Connecting every bit of knowledge: \\
The Structure of Wikipedia's first link network
}

@@

\begin{abstract}
  \protect
Apples, porcupines, and the most obscure Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in an article, we connect entries to form a directed network: Wikipedia's First Link Network. 
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many 
inventions, places, people, objects, and events
are related and organized.  
We algorithmically parse all 4.7 million articles to construct a map of Wikipedia's First Link Network. 
We traverse every possible path through the network, 
measuring the accumulation of first links, path lengths, basins, cycles, and the influence a particular article exerts in shaping the 
network.
We discover scale-free distributions describe path length, 
accumulation, and influence. Far from dispersed, first links disproportionately accumulate 
at a few articles---flowing from specific to general and culminating around fundamental notions such as
Community, State, and Science. Even fewer articles most influence the network's structure with Philosophy 
shaping more paths than any other article by two orders of magnitude. 
Curiously, we also observe a gravitation towards topical articles such as 
Health Care and Fossil Fuel. 
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

Wikipedia is a towering achievement of the modern era. 
At no point in history has a larger or more meticulously indexed collection of human knowledge 
existed.
Wikipedia contains 37 million articles in 283 languages, 
with coverage spanning everything from little known ancient battles to the latest pharmaceutical drugs 
\cite{drugs, stats}.
All this information does not lie dormant. 
Wikipedia is the sixth most visited
site in the world, surpassing 18 billion page views and 26 million edits in a single month
\cite{wiki_edits, wiki_views}.

Wikipedia has naturally become the object of many studies. 
Researchers have examined the cultural dynamics among editors
\cite{editors},
the accuracy of the content relative to traditional encyclopedias
\cite{accuracy1, accuracy2},
the topics covered 
\cite{coverage},
and bias against portions of the population
\cite{bias_women}.
Wikipedia's content has also proven a powerful tool. 
Researchers have used Wikipedia to identify missing dictionary entries
\cite{missing_entries},
cluster short text
\cite{clustering},
compute semantic relatedness
\cite{semantic_relatedness},
and disambiguate meaning \cite{disambiguating}.

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
chose
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.

@@

By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the topic of one article to that of another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{graphics/Train.pdf}  
  \caption{
    \textbf{First Link Path For ``Train.''}
    We follow the first link to another Wikipedia article
    in the main body of the article---the area
    inside the green rectangle, which excludes 
    side bar elements, the navigation bar and title. 
    In this example, the first link to another Wikipedia article is ``Rail Transport.'' 
    We can again select the first link on the ``Rail Transport''
    article, repeating the process to 
    form a path of first links.
  }
  \label{fig:Train First Links}
\end{figure*}

@@

Our goal is to study the structure of the FLN for insight into how the information on Wikipedia is organized and related.
We develop metrics to capture 
the dominant features of the FLN's structure.
We measure dynamics of the FLN as a flow, quantifying 
aspects of the FLN from the accumulation of first links around articles 
to the influence an article exerts in shaping the FLN.
Together with cycles, in-degree, depth, and the content of the articles, 
we build our analysis of the relations among the ideas in Wikipedia.

@@

An essential feature of a directed network's structure is the degree distribution 
\cite{newman}. 
The degree distribution has been used to study many phenomena from disease outbreak 
\cite{disease} 
to the dynamics of social networks 
\cite{social_nets}.
The in-degree distribution in the FLN describes how many first links point to a 
particular article. 
We measure the in-degree in the FLN to gauge the number of articles referencing a particular article.
Articles with zero in-degree have no references---they are outer leaves in the FLN. 
Similar to PageRank, the in-degree 
provides a way to rank the articles in Wikipedia, allowing us to see 
whether authors tend to reference all articles equally or only a few.

Our greater interest is lies in the richer dynamics of how many articles fit together.
Inspired by the claim that the majority of first links lead to
``Philosophy''---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{mat_blog, Ilmari_first_links, xkcd}
---
we holistically study how the FLN  yields a flow of connections.
We construct a path through the FLN with each article 
leading to the next through the first link. 
For example ``Train'' has a first link to 
``Rail Transport,''
 which is itself
an article with a first link to 
``Conveyance of Passengers and Goods,''
and so on.
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Grounded in the dynamics of river networks
\cite{geo_basins},
we characterize features of the FLN as a flow through first links.
Previous studies have used flow to characterize the structure of river networks
\cite{dodds} and describe the organization of food systems through transportation networks
\cite{food_webs}.
With paths to mark flow through the FLN, we develop metrics to 
gauge the accumulation of references, 
the length of the path relating articles, and the influence a particular
article exerts in shaping the flow of links through the FLN.

@@

We can characterize the paths in the FLN as a matrix with each column corresponding to a path. In our sample network 
(Fig.~\ref{fig:Traversal Visits}), the path starting at article A is the 
first column in the traversal visits matrix. 
An entry of $1$ indicates the path contains a given article and 
$0$ indicates the path does not.
To compute the number of traversal visits for an article, we sum the corresponding row in our matrix.
The traversal visits matrix for 
our Wikipedia dataset consists of 121 million entries to account for each path through the FLN.

By measuring the number of first links between
two articles, we obtain an additional piece of information we call 
{\it path length.} We can compute the length of a path by summing 
along a column in our traversal visits matrix. 
In our Wikipedia dataset, the sum of all path lengths is 232 million first links.
The path length describes how closely related topics are.
The greater the number of first links separating two articles, the greater the number of ideas needed to relate their corresponding ideas. Although
``Train'' is related to ``Economics'' for example, there are several articles
bridging the connection:
``Train'' is more specifically related to transportation, whose object
is often goods. Goods are one of the fundamental objects of study in Economics.
Described in links, this relationship is captured by a relatedness of $4$ 
first links ultimately connecting ``Train'' to ``Economics.''

@@

While traversal visits measure accumulation, each article's first link also 
influences the shape of the FLN. 
At a large point of accumulation, a single article's first link 
can exert great influence over the shape of the FLN by directing many
references on a particular path. To distinguish between an article 
that simply happened to fall within a cycle from an article funneling 
many first links, we develop a second metric called {\it traversal funnels}.

To measure traversal funnels, we traverse the FLN in the same manner as we 
did for traversal visits, but end a path once we enter a cycle.
We are then able to distinguish between an article related to many other ideas
only by virtue of its place in a cycle, from an article exerting influence over where the first links flow. 
An article with a large number of traversal funnels directs many references
into a particular cycle. In our sample network 
(see Fig.~\ref{fig:Traversal Funnels}) article C 
directs the flow of links towards the 3-cycle, while articles A and B are 
recipients of the flow---without exerting any direct influence themselves.

@@

\section{Appendix}

>>>>

\title{\protect
Connecting every bit of knowledge: \\
The structure of Wikipedia's first link network
}

@@

\begin{abstract}
  \protect
Apples, porcupines, and the most obscure Bob Dylan song---is every topic a few clicks from Philosophy? 
Within Wikipedia, the surprising answer is yes: nearly all 
paths lead to Philosophy.
Wikipedia is the largest, most meticulously indexed collection of human knowledge ever amassed. 
More than information about a topic, Wikipedia is a web of naturally emerging relationships.  
By following the first link in each article, we algorithmically construct a directed network of 
all 4.7 million articles: Wikipedia's First Link Network.
Here, we study the English edition of Wikipedia's First Link Network for insight into how the many 
inventions, places, people, objects, and events are related and organized.  
We traverse every path, measuring the accumulation of first links, path lengths, basins, cycles, 
and the influence each article exerts in shaping the network.
We discover scale-free distributions describe path length, 
accumulation, and influence. Far from dispersed, first links disproportionately accumulate 
at a few articles---flowing from specific to general and culminating around fundamental notions such as
Community, State, and Science. 
Philosophy shapes more paths than any other article by two orders of magnitude. 
Curiously, we also observe a gravitation towards topical articles such as 
Health Care and Fossil Fuel. 
These findings enrich our view of the connections and structure of
Wikipedia's ever growing store of knowledge.

@@

Wikipedia is a towering achievement of the modern era. 
At no point in history has a larger or more meticulously indexed collection of human knowledge 
existed.
Wikipedia contains 37 million articles in 283 languages, 
with coverage spanning everything from little known ancient battles to the latest pharmaceutical drugs 
\cite{drugs, stats}.
Demonstrating its relevance to modern inquiry,
Wikipedia is the sixth most visited
site in the world, surpassing 18 billion page views and 26 million edits in a single month
\cite{wiki_edits, wiki_views}.

Wikipedia has naturally become the object of many studies. 
Researchers have examined the cultural dynamics among editors
\cite{editors},
the accuracy of the content relative to traditional encyclopedias
\cite{accuracy1, accuracy2},
the topics covered 
\cite{coverage},
and bias against portions of the population
\cite{bias_women}.
Wikipedia's content has also proven to be a powerful tool. 
Researchers have used Wikipedia to identify missing dictionary entries
\cite{missing_entries},
cluster short text
\cite{clustering},
compute semantic relatedness
\cite{semantic_relatedness},
and disambiguate meaning \cite{disambiguating}.

@@

The authors of a Wikipedia article choose where and whether to include a 
reference to another Wikipedia article
in the HTML markup.
For example, the authors of the 
\href{https://en.wikipedia.org/wiki/Train}{``Train''}
article
chose
``Amtrak's Acela Express,'' ``steam,'' and ``head-end power'' 
among others as relevant articles to reference in describing ``Train''
\cite{wiki_train}
.

@@

By following the first hyperlink to another article in the English edition of
Wikipedia as of November, 2014, we connect the topic of one article to that of another, ultimately forming a directed network: 
{\it Wikipedia's First Link Network} (FLN).
For methodological details, see section~\ref{Appendix}.

\begin{figure*}[tp!]
  \includegraphics[width=\textwidth]{graphics/Train.pdf}  
  \caption{
    \textbf{First Link Path For ``Train.''}
    We follow the first link  to another Wikipedia article
    in the main body of the article---the area
    inside the green rectangle, which excludes 
    side bar elements, the navigation bar and title; the first link is circled in red. 
    In this example, the first link to another Wikipedia article is ``Rail Transport.'' 
    We can again select the first link on the ``Rail Transport''
    article, repeating the process to 
    form a path of first links. After 11 links, we arrive at ``Philosophy.''
  }
  \label{fig:Train First Links}
\end{figure*}

@@

Our goal is to study the structure of the FLN for insight into how the information on Wikipedia is organized and related.
We develop metrics to capture 
the dominant features of the FLN's structure.
We measure dynamics of the FLN as a flow, quantifying 
the accumulation of first links around articles 
and the influence an article exerts in shaping the FLN.
Together with cycles, in-degree, depth, and the content of the articles, 
we build our analysis of the relations among the ideas in Wikipedia.

@@

An essential feature of a directed network's structure is the degree distribution 
\cite{newman}. 
The degree distribution has been used to study many phenomena from disease outbreak 
\cite{disease} 
to the dynamics of social networks 
\cite{social_nets}.
The in-degree distribution in the FLN describes how many first links point to a 
particular article. 
Articles with zero in-degree have no references---they are outer leaves in the FLN. 
The in-degree 
provides a way to rank the articles in Wikipedia by the number of first link references.

Inspired by the claim that the majority of first links lead to
``Philosophy''---popularized by an 
\href{https://xkcd.com/903/}{xkcd}
comic and subsequently
discussed in blog posts 
\cite{xkcd, mat_blog, Ilmari_first_links}
---
we holistically study how the FLN  yields a flow of connections.
For example ``Train'' has a first link to 
``Rail Transport,''
 which is itself
an article with a first link to 
``Conveyance of Passengers and Goods,''
and so on.
As shown in Fig.~\ref{fig:Train First Links}, the path starting at ``Train'' 
contains ``Goods,'' ``Economics,'' ``Social Science,''
leading ultimately to ``Philosophy''.

@@

Starting at each article, we construct a path through the FLN, 
to map the flow of connections among articles.
The method is order agnostic with respect to which articles are selected first. As long as each article is selected eventually, the resulting metrics are equivalent.
Previous studies have used flow to characterize the structure of river networks
\cite{geo_basins, dodds} and describe the organization of food systems through transportation networks
\cite{food_webs}.
With paths to mark flow through the FLN, we develop metrics to 
gauge the accumulation of references, 
the length of the path relating articles, and the influence a particular
article exerts in shaping the flow of links through the FLN.

@@

We can characterize the paths in the FLN as a matrix with each column corresponding to a path. In our sample network 
(Fig.~\ref{fig:Traversal Visits}), the path starting at article A is the 
first column in the traversal visits matrix. 
An entry of $1$ indicates the path contains a given article and 
$0$ indicates the path does not.
To compute the number of traversal visits for an article, we sum the corresponding row in our matrix.
The traversal visits matrix for 
our Wikipedia dataset consists of 121 million entries encoding each path out of the more than 
googol possible paths ($4.7*10^6*2^{4.7*10^6}$) through the FLN .

By measuring the number of first links between
two articles, we obtain an additional piece of information we call 
{\it path length.} We can compute the length of a path by summing 
along a column in our traversal visits matrix. 
In our Wikipedia dataset, the sum of all path lengths is 232 million first links.
The path length describes how closely related topics are.
Although ``Train'' is related to ``Economics'' for example, there are several articles
bridging the connection:
``Train'' is more specifically related to transportation, whose object
is often goods. Goods are one of the fundamental objects of study in Economics.
Described in links, this relationship is captured by a relatedness of $4$ 
first links ultimately connecting ``Train'' to ``Economics.''

@@

While traversal visits measure accumulation, each article's first link also 
influences the shape of the FLN. 
At a point of accumulation, a single article's first link 
can exert great influence over the shape of the FLN by directing many
references on a particular path. To distinguish between an article 
that simply happened to fall within a cycle from an article funneling 
many first links, we develop a second metric called {\it traversal funnels}.

To measure traversal funnels, we traverse the FLN in the same manner as we 
did for traversal visits, but end a path once we enter a cycle.
We are then able to distinguish between an article related to many other ideas
only by virtue of its place in a cycle, from an article exerting influence over where the first links flow. 
An article with a large number of traversal funnels directs many references
into a particular cycle. In our sample network 
(see Fig.~\ref{fig:Traversal Funnels}) article C 
directs the flow of links towards the 3-cycle, while articles A and B are 
recipients of the flow---without exerting direct influence themselves.

@@

\section{Appendix}
\label{Appendix}
